A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-07-24 doc/large_strain.tex

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches \cite{kronbichler12, kronbichler2017fast} where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

For the performance of the matrix-free operator, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.
%
In order to obtain gradients with respect to the spatial configuration,
that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem derived with a Lagrangian formulation,
we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
  cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ \;
        evaluate the following tensors at each quadrature point using sum factorization :\\
        \enskip $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
        \enskip $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        \enskip $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ \tcp*{2nd order}
          evaluate $J = \rm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \rm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
      evaluate queued contractions using sum factorization \;
      distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache scalar quantities only}
  \label{alg:mf_scalar}
\end{algorithm}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        extract local vector values on this element $\gz x_K$ \;
        evaluate the following tensors at each quadrature point using sum factorization :\\
        \enskip$\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        \enskip$\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \rm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
        evaluate queued contractions using sum factorization \;
        distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
  \label{alg:mf_tensor2}
\end{algorithm}

@@

The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} from the Epetra package \cite{Heroux2005b} sparse data structure and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is available in \texttt{deal.II}.

@@

Aligned with the approach taken in similar previous investigations, compare e.g. \cite{kronbichler2017fast},
we limit ourselves to comparisons made on a node level in this study.
Examinations performed in this manner should already indicate the differences between the matrix-free and matrix-based approaches, and give some preliminary understanding of the scalability potential of the numerical implementation without confounding factors related to data exchange through interconnects via MPI.
The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial order finite element.
Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound.

We conduct two performance studies using the described discretization, the first being for only matrix-vector multiplication, and the second for the preconditioned iterative linear solver.
The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library version 12.12.1.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fifth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.
On the coarsest level, we use the Chebyshev preconditioner as a solver \cite{Varga2009} to reduce the residual by three orders of magnitude.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Emmy_RRZE_timing2d.eps}
      \caption{matrix-vector product (2D)}
      \label{fig:benchmark_miehe_Emmy_vmult2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Emmy_RRZE_timing3d.eps}
    \caption{matrix-vector product (3D)}
    \label{fig:benchmark_miehe_Emmy_vmult3}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Emmy_RRZE_memory2d.eps}
      \caption{memory consumption (2D)}
      \label{fig:benchmark_miehe_Emmy_memory2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Emmy_RRZE_memory3d.eps}
    \caption{memory consumption (3D)}
    \label{fig:benchmark_miehe_Emmy_memory3}
  \end{subfigure}
  \caption{Emmy cluster, RRZE, matrix-vector multiplication.}%
  \label{fig:benchmark_miehe_Emmy}
\end{figure}

@@

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
We are interested in the wall-clock time (average over 10 runs per each linear solver step) per DoF as a first metric.
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
as the matrix becomes more and more dense.
The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces \cite{Arndt2017a} to capture the geometry as well as
the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
Both factors increase the memory consumption and make element operations more expensive.
Nonetheless,
the matrix-free implementation is faster than the matrix-based counterpart already for bi-cubic Lagrangian basis in 2D and tri-quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference in terms of the wall-clock time per DoF between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.

@@

When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
we observe a marginal increase in computational intensity and almost unchanged performance.
This coincides with our expectation that the latter algorithm requires transferring fewer bytes to perform operations on a single quadrature point
(compare line 9 in Algorithm \ref{alg:mf_tensor4} that loads a symmetric rank-4 tensor to the line 9 in Algorithm \ref{alg:mf_tensor2} that evaluates the
same contraction by loading two scalars $c_1$ and $c_2$).
On the other hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
This is not surprising as at each quadrature point we need to re-evaluate the Kirchoff stress $\gz \tau$.
Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
second-order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
however the ``scalar'' algorithm performs a large number of repetitive operations that are
independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.

@@

In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
(i) `read / write' denotes reading and writing from/to a global vector (lines 2 and 13);
(ii) `sum factorization' denotes application of sum factorization techniques (lines 4, 5 and 12);
(iii) `quadrature loop' denotes operations performed on each quadrature (lines \mbox{6 -- 11}).
Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
global refinement in 2D and no global mesh refinement in 3D.
It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that the
`read / write' steps take a negligible amount of time on each element.
For higher order bases, the majority of time is spent within the quadrature loop.
When analyzing the roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
In this case the, quadrature loop consists of, essentially, a
single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).

>>>>

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches \cite{kronbichler12, kronbichler2017fast} where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form at one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$, \todo{Should we include the number of elements in the total mat-vec?}
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

For the performance of the matrix-free operator evaluation, it is crucial to find a good intermediate point between precomputing everything (caching) and evaluating all the quantities on-the-fly.
%
In order to obtain gradients with respect to the spatial configuration,
that appear in the tangent operator \eqref{eq:algebraic_tangent} of the finite-strain elasticity problem derived with a Lagrangian formulation,
we used the \textit{MappingQEulerian} class from the \texttt{deal.II} \cite{dealII90} library.
In addition to the standard mapping from the reference (isoparametric) element to the element in real space, this class computes the mapping to the spatial configuration given a displacement field.
%
Due to the specifics of matrix-free operator evaluation implemented in \texttt{deal.II}, the integration is eventually performed over the spatial configuration.
Therefore, we have to additionally divide by the Jacobian $J$ of the deformation map at a given quadrature point to express the integral in the referential configuration.

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$, current FE solution $\overline{\gz u^h}$,
  cached $c_1 := \mu - 2 \lambda \log(J)$ for each cell and quadrature point}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        extract local vector values on this element $\overline{\gz u^h}_K$, $\gz x_K$ \;
        evaluate the following tensors at each quadrature point using sum factorization :\\
        \enskip $\Grad \, \overline{\gz u^h}_K$ \tcp*{2nd order}
        \enskip $\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        \enskip $\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\gz F = \gz I + \Grad \, \overline{\gz u^h}$ \tcp*{2nd order}
          evaluate $J = \textrm{det}(\gz F)$ \tcp*{scalar}
          evaluate $\gz b = \gz F \cdot \gz F^T$ \tcp*{2nd order symmetric}
          evaluate $\gz \tau = \mu \gz b - c_1 \gz I$ \tcp*{2nd order symmetric}
          evaluate $\boldsymbol{\mathcal{G}} \gz g := \gz g \cdot \gz \tau/J$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$\;
          evaluate $\boldsymbol{\mathcal{C}}\gz g_s := \left[2 c_1 \gz g_s + 2 \lambda\, \textrm{tr}( \gz g_s) \gz I \right]/J$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ for contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}}\gz g_s$ \;
        }
      evaluate queued contractions using sum factorization \;
      distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache scalar quantities only}
  \label{alg:mf_scalar}
\end{algorithm}

@@

\begin{algorithm}[!ht]
  \SetKwInOut{Input}{Given}
  \SetKwInOut{Output}{Return}
  \Input{Source FE vector $\gz x$,
  cached
  $c_1 := 2\left[\mu - 2 \lambda \log(J)\right]/J$,
  $c_2 := 2\lambda/J$ and $\gz \tau/J$ for each cell and quadrature point,
  evaluated based on $\overline{\gz u^h}$}
  \Output{action of the FE tangent operator \eqref{eq:algebraic_tangent} on $\gz x$}
  \ForEach{ element $K \in \Omega^h$ }{
        extract local vector values on this element $\gz x_K$ \;
        evaluate the following tensors at each quadrature point using sum factorization :\\
        \enskip$\gz g_s := \overline{\grad^s} \gz x_K$ \tcp*{2nd order symmetric}
        \enskip$\gz g := \overline{\grad} \gz x_K$ \tcp*{2nd order}
        \ForEach{quadrature point $q$ on $K$}{
          evaluate $\boldsymbol{\mathcal{G}}\gz g := \gz g \cdot \left[\gz \tau/J \right]$ \tcp*{2nd order}
          queue $\boldsymbol{\mathcal{G}} \gz g$ for contraction $\overline{\grad} \gz N_i : \boldsymbol{\mathcal{G}} \gz g$ \;
          evaluate $\boldsymbol{\mathcal{C}} \gz g_s := c_1 \gz g_s + c_2 \, \textrm{tr}(\gz g_s) \gz I$ \tcp*{2nd order symmetric}
          queue $\boldsymbol{\mathcal{C}} \gz g_s$ contraction $\overline{\grad^s} \gz N_i : \boldsymbol{\mathcal{C}} \gz g_s $ \;
        }
        evaluate queued contractions using sum factorization \;
        distribute results to the destination vector
  }
  \caption{Matrix-free tangent operator: cache second-order Kirchhoff stress $\gz \tau$ and thereby avoid the need to evaluate referential quantities like $\gz F$ at when executed.}
  \label{alg:mf_tensor2}
\end{algorithm}

@@

The discrete tangent operator described by \eqref{eq:algebraic_tangent} can be implemented in one of two ways, namely through a classical matrix-based approach or, alternatively, a matrix-free approach.
For the matrix-based approach adopted in this study, the Trilinos \cite{Heroux2005} library was leveraged;
the system tangent matrix is stored in a distributed \textit{Epetra\_FECrsMatrix} sparse data structure from the Epetra package \cite{Heroux2005b} and the linear solver was preconditioned with an algebraic multigrid (AMG) preconditioner from the ML package \cite{Gee2006a}.
This is the most highly performant matrix-based approach using the Trilinos framework that, to date, is available in \texttt{deal.II}.

@@

Aligned with the approach taken in similar previous investigations, compare e.g. \cite{kronbichler2017fast},
we limit ourselves to comparisons made on a node level in this study. Examinations performed in this manner are most representative for the differences between the matrix-free and matrix-based approaches because problems of sufficient size are dominated by the local node-level performance with multigrid solvers. Multigrid solvers are well-known to be scalable given sufficiently small coarse-grid problems even to the largest supercomputers \cite{gholami2016,ibeid2018} as the communication is mostly between nearest neighbors.
The benchmark calculations are performed for 2D and 3D problems with various combinations of polynomial degrees and number of global refinements, as is stated in Tables \ref{tab:input_parameters_2d} and \ref{tab:input_parameters_3d}.
We chose a problem size around $10^6$ DoFs for each test case as this was an upper limit due to the memory requirements of the matrix-based approach with a high polynomial order finite element.
Even so, the problem size is large enough so that the sparse matrix (in excess of $1$ Gb when using the lowest polynomial order finite element discretization) will not fit into the L3 cache (on the order of tens of Mb in size), and therefore we can observe that the matrix-based approach is memory bound.

We conduct two performance studies using the described discretization, the first being for matrix-vector multiplication only, and the second for the preconditioned iterative linear solver.
The results of the matrix-free approach with a geometric multigrid preconditioner are compared to the matrix-based approach in conjunction with the algebraic multigrid (AMG) preconditioner using the package ML \cite{Gee2006a} of the Trilinos \cite{Heroux2005} library version 12.12.1.
The aggregation threshold for the AMG preconditioner is taken as $10^{-4}$.
The geometric multigrid algorithm uses a V-cycle with two pre- and post-smoothing steps using Chebyshev polynomials \cite{Varga2009} of fifth degree, provided
by the \texttt{deal.II} library via the \textit{PreconditionChebyshev} class.
The largest eigenvalue $\lambda_{\rm{max}}$ of the level operators is estimated from $30$ iterations of the CG solver and the smoother
is set to target the range $[0.6\lambda_{\rm{max}}, 1.2\lambda_{\rm{max}}]$.\todo{Does this work well (verified in code)? I would use $0.1\lambda_{\rm{max}}$, i.e., smoothing range 12.}
On the coarsest level, we use the Chebyshev preconditioner as a solver \cite{Varga2009} to reduce the residual by three orders of magnitude.
Since this smoother only uses diagonal elements of the operator it is readily compatible with the matrix-free approach while most default smoothers rely on an explicit matrix form.

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Emmy_RRZE_timing2d.eps}
      \caption{matrix-vector product (2D)}
      \label{fig:benchmark_miehe_Emmy_vmult2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Emmy_RRZE_timing3d.eps}
    \caption{matrix-vector product (3D)}
    \label{fig:benchmark_miehe_Emmy_vmult3}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
      \centering
      \includegraphics[width=\textwidth]{Emmy_RRZE_memory2d.eps}
      \caption{memory consumption (2D)}
      \label{fig:benchmark_miehe_Emmy_memory2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{Emmy_RRZE_memory3d.eps}
    \caption{memory consumption (3D)}
    \label{fig:benchmark_miehe_Emmy_memory3}
  \end{subfigure}
  \caption{Emmy cluster, 20 cores, RRZE, matrix-vector multiplication.}%
  \label{fig:benchmark_miehe_Emmy}
\end{figure}

@@

\todo{Is the IWR result also run with 20 MPI ranks? How are MPI ranks pinned to the 80 physical cores?}

Figures \ref{fig:benchmark_miehe_Emmy} and \ref{fig:benchmark_miehe_IWR} show the results of the matrix-vector multiplication for the considered finite-strain hyperelastic benchmark problem as performed on the two clusters.
We are interested in the wall-clock time (average over 10 runs for each linear solver step) per DoF as a first metric.
As expected, the matrix-vector multiplication becomes very expensive for higher order discretization for sparse matrix-based approaches,
as the matrix becomes more and more dense.
The performance results for finite-strain elasticity operators considered in this study are qualitatively similar to those reported for the Laplace operator in \cite{kronbichler12}.
Note that we consider only non-Cartesian meshes together with additional mappings (namely, manifold descriptions of the boundaries and interfaces \cite{dealII90} to capture the geometry as well as
the linearization-related mapping required for evaluation of gradients with respect to the deformed configuration).
Both factors increase the memory consumption and make element operations more expensive.
Nonetheless,
the matrix-free implementation is faster than the matrix-based counterpart already for bi-cubic Lagrangian basis in 2D and tri-quadratic in 3D.
Figures \ref{fig:benchmark_miehe_Emmy_vmult2}, \ref{fig:benchmark_miehe_Emmy_vmult3}, \ref{fig:benchmark_miehe_IWR_vmult2} and \ref{fig:benchmark_miehe_IWR_vmult3} clearly show the influence of the different caching strategies,
described in Algorithms \ref{alg:mf_scalar}, \ref{alg:mf_tensor2} and \ref{alg:mf_tensor4}.
Two conclusions can be drawn from these results:
The scalar caching implementation is the most time consuming of the three, as at each step we additionally evaluate the gradient of the displacement field in the referential configuration that is required to evaluate the Kirchhoff stress $\gz \tau$.
There is little difference in terms of the wall-clock time per DoF between the approach where the material part of the fourth-order spatial tangent stiffness tensor is cached (``tensor4'') and the one where we only cache the second-order Kirchhoff tensor and utilize the chosen hyperelastic constitutive equation to efficiently implement the action of the material part of the fourth-order spatial tangent stiffness tensor on the second-order symmetric tensor (``tensor2'').
This indicates that the time savings we get from the latter approach are insignificant within the complete matrix-free operator implementation.
Consequently, we can apply the matrix-free approach to any finite-strain material model by caching the material part of the fourth-order spatial tangent stiffness tensor in addition to the Kirchhoff stress and expect this to be faster than the matrix-based implementation.
It is therefore feasible to define highly performant generic operators that are independent of any applied constitutive laws,
as long as the resulting tangent operator is expressed via \eqref{eq:algebraic_tangent}.

@@

When comparing the data for the ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}) caching strategy,
we observe a marginal increase in computational intensity and almost unchanged performance.
This coincides with our expectation that the latter algorithm requires transferring fewer bytes to perform operations on a single quadrature point
(compare line 9 in Algorithm \ref{alg:mf_tensor4} that loads a symmetric rank-4 tensor to the line 9 in Algorithm \ref{alg:mf_tensor2} that evaluates the
same contraction by loading two scalars $c_1$ and $c_2$).
On the other hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
This is not surprising as at each quadrature point we need to re-evaluate the Kirchoff stress $\gz \tau$.
Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
second-order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
As a result both algorithms will have comparable memory requirements (also confirmed by Figures \ref{fig:benchmark_miehe_Emmy_memory2}, \ref{fig:benchmark_miehe_Emmy_memory3}, \ref{fig:benchmark_miehe_IWR_memory2} and \ref{fig:benchmark_miehe_IWR_memory3}),
however the ``scalar'' algorithm performs a large number of repetitive operations that are
independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.

@@

In order to have a better understanding of the behavior of the most promising ``tensor4'' implementation,
we collect its performance data and computing times for various stages inside the element loop (see Algorithm \ref{alg:mf_tensor4}):
(i) `read / write' denotes reading and writing from/to a global vector (lines 2 and 13);
(ii) `sum factorization' denotes application of sum factorization techniques (lines 4, 5 and 12);
(iii) `quadrature loop' denotes operations performed at each quadrature (lines \mbox{6 -- 11}).
Given that the measurements are now down within the cell loop, we collected results on smaller meshes by adopting only one
global refinement in 2D and no global mesh refinement in 3D.
It is clear from Figure \ref{fig:breakdown_stackedbar_2d_tensor4} and  \ref{fig:breakdown_stackedbar_3d_tensor4} that the
`read / write' steps take a negligible amount of time on each element.
For higher order bases, the majority of time is spent within the quadrature loop.
When analyzing the roofline data (Figure \ref{fig:roofline_2d_tensor4} and \ref{fig:roofline_3d_tensor4}),
we observe that this part of the algorithm demonstrates rather low performance, which in our opinion is the source of the
suboptimal performance of the whole matrix-free matrix-vector product (see Figure \ref{fig:roofline}).
In this case the, quadrature loop consists of, essentially, a
single contraction between two second order tensors (step 7 in Algorithm \ref{alg:mf_tensor4}) and a
double contraction between a symmetric second order tensor and a fourth order tensor (step 9 in Algorithm \ref{alg:mf_tensor4}).
